---
title: "Graphical Models Homework - 4"
author: "Bassel MASRI"
date: "12/21/2020"
output:
html_notebook: default
---

3. Implementing Meinshausen-Buhlmann and the Glasso algorithms on the Ecoli network.

```{r, warning=FALSE}
#-- clean environment and load library
rm(list=ls())
library(huge)
library(igraphdata)
library(igraph)
library(sand)
```

```{r}
#-- Meinshausen-Buhlmann algorithm
huge_mb.out = huge(Ecoli.expr, method = "mb", nlambda=NULL,lambda.min.ratio = NULL)

#-- optimal lambda RIC
huge_mb.opt_RIC = huge.select(huge_mb.out,criterion="ric")
huge_mb.opt_RIC$opt.lambda
g1 = graph.adjacency(huge_mb.opt_RIC$refit,"undirected")

#-- optimal lambda Stability selection criterion
huge_mb.opt_stars = huge.select(huge_mb.out,criterion="stars")
huge_mb.opt_stars$opt.lambda
g2 = graph.adjacency(huge_mb.opt_stars$refit,"undirected")


#-- GLASSO
huge_glasso.out = huge(Ecoli.expr,method = "glasso", nlambda=NULL,lambda.min.ratio = NULL)

#-- optimal lambda RIC
huge_glasso.opt_RIC = huge.select(huge_glasso.out,criterion="ric")
huge_glasso.opt_RIC$opt.lambda
g3 = graph.adjacency(huge_glasso.opt_RIC$refit,"undirected")

#-- optimal lambda Stability selection criterion
huge_glasso.opt_stars = huge.select(huge_glasso.out,criterion="stars")
huge_glasso.opt_stars$opt.lambda
g4 = graph.adjacency(huge_glasso.opt_stars$refit,"undirected")
```

4. Comparing networks

```{r, fig.align='center', fig.width=6, fig.height=6}
#-- get intersection
inter = graph.intersection(g2,g4)
plot(inter, vertex.color = "blue4", 
     vertex.size = 6, 
     edge.arrow.size=.4,
     vertex.label=NA, 
     edge.color = "grey40")
```

```{r, fig.align='center', fig.width=12, fig.height=6}
#-- compute degree distribution of each network
par(mfrow = c(1,2))

hist(degree(g2), main = "Degree distribution of Meinshausen-Buhlmann \nmodel with optimal lambda",
     xlab = "Degree",
     ylab = "Frequency",
     col = "grey",
     breaks=20)

hist(degree(g4), main = "Degree distribution of GLASSO \nwith optimal lambda",
     xlab = "Degree",
     ylab = "Frequency",
     col = "grey",
     breaks=20)
```

```{r}
#-- number of edges of the created networks
E(g2)
E(g4)
```

We notice a significant difference in the degree distribution of both models. The Meinshausen-Buhlmann seems to follow a more gaussian curve whereas the GLASSO seems to resemble a decaying power law distribution. 

We also seem to have different number of edges. For the Meinshausen-Buhlmann model we have 623 edges whereas the GLASSO only contains 554 edges.

5. Comparing with the Ecoli network

```{r}
#-- construct the ecoli graph
g.regDB=graph.adjacency(regDB.adj,"undirected")

#-- get intersection with Meinshausen-Buhlmann (with optimal lambda)
graph.intersection(g.regDB, g2, byname = FALSE)

#-- get intersection with GLASSO (with optimal lambda)
graph.intersection(g.regDB, g4, byname = FALSE)
```

We notice that the Meinshausen-Buhlmann was able to capture 19 genes in the original Ecoli network whereas the GLASSO model was able to capture only 9.